MCMC convergence diagnostics`

### 2023-01-12 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Clean up MCMC convergence notebook

---
title: "MCMC convergence diagnostics`"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background**

  **Task** 
---

```{r}
out <- readRDS("depends/out.rds")

mcmc <- out$mcmc

bayesplot::color_scheme_set("viridisA")
ggplot2::theme_set(theme_minimal())

#' Rhat
#' Looking for values less than 1.1 here
rhats <- bayesplot::rhat(mcmc)
bayesplot::mcmc_rhat(rhats)
(big_rhats <- rhats[rhats > 1.1])
length(big_rhats) / length(rhats)

#' ESS ratio
#' Worry about values less than 0.1 here... which they all are :clown:
ratios <- bayesplot::neff_ratio(mcmc)
bayesplot::mcmc_neff(ratios)

#' Autocorrelation
#' High values of autocorrelation in the chains
bayesplot::mcmc_acf(mcmc, pars = vars(starts_with("beta")))

#' Univariate traceplots

#' Assess these visually
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("beta")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("logit")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_sigma")))

#' Prevalence model
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_rho_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_a[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_as[")))

#' ART model
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_alpha_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_a[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_as[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_xa[")))

#' Other
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_lambda_x[")))

#' ANC model
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_anc_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_anc_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_or_gamma["))) #' N.B. these are from the ANC attendance model

#' Pairs plots
#' Prior suspicion (from Jeff, Tim, Rachel) that the ART attendance model is unidentifiable
#' Let's have a look at the pairs plot for neighbouring districts and the log_or_gamma parameter
nb <- area_merged %>%
  filter(area_level == max(area_level)) %>%
  bsae::sf_to_nb()

@@

area_merged %>%
  filter(area_level == max(area_level)) %>%
  print(n = Inf)
neighbours_pairs_plot("log_or_gamma", 5) #' Nkhata Bay and neighbours
neighbours_pairs_plot("log_or_gamma", 26) #' Blantyre and neighbours

#' NUTS specific assessment
np <- bayesplot::nuts_params(mcmc)

#' Number of divergent transitions
#' In this instance it's zero, so no need to do further divergent transition investigation
np %>%
  filter(Parameter == "divergent__") %>%
  summarise(n_divergent = sum(Value))

bayesplot::mcmc_nuts_divergence(np, bayesplot::log_posterior(mcmc))

#' Energy plots
#' Ideally these two histograms would be the same (Betancourt 2017)
#' The histograms are quite different, in a way suggesting the chains may not be
#' fully exploring the tails of the target distribution
bayesplot::mcmc_nuts_energy(np)
```

>>>>

---
title: "MCMC convergence diagnostics`"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
  **Background** In the report `naomi-simple_fit` with parameter `tmbstan = TRUE`, we used the NUTS algorithm to perform MCMC inference for the simplified Naomi model.

  **Task** Here we assess whether or not the results of the MCMC are suitable using a range of diagnostic tools.
---

We start by obtaining results from the latest version of `naomi-simple_fit` with `tmbstan = TRUE`.

```{r}
out <- readRDS("depends/out.rds")
mcmc <- out$mcmc
```

This MCMC took `r as.numeric(out$time)` minutes to run

```{r}
bayesplot::color_scheme_set("viridisA")
ggplot2::theme_set(theme_minimal())
```

# $\hat R$

We are looking for values of $\hat R$ less than 1.1 here.

```{r}
rhats <- bayesplot::rhat(mcmc)
bayesplot::mcmc_rhat(rhats)
(big_rhats <- rhats[rhats > 1.1])
length(big_rhats) / length(rhats)
```

# ESS ratio

Reasonable to be worried about values less than 0.1 here.

```{r}
ratios <- bayesplot::neff_ratio(mcmc)
bayesplot::mcmc_neff(ratios)
```

# Autocorrelation

How much autocorrelation is there in the chains?

```{r}
bayesplot::mcmc_acf(mcmc, pars = vars(starts_with("beta")))
```

# Univariate traceplots

```{r}
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("beta")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("logit")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_sigma")))
```

## Prevalence model

```{r}
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_rho_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_a[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_rho_as[")))
```

## ART model

```{r}
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("us_alpha_xs[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_a[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_as[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("u_alpha_xa[")))
```

## Other

```{r}
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_lambda_x[")))
```

## ANC model

```{r}
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_anc_rho_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("ui_anc_alpha_x[")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_or_gamma["))) #' N.B. these are from the ANC attendance model
```

# Pairs plots

There is a prior suspicion (from Jeff, Tim, Rachel) that the ART attendance model is unidentifiable.
Let's have a look at the pairs plot for neighbouring districts and the `log_or_gamma` parameter.

```{r}
nb <- area_merged %>%
  filter(area_level == max(area_level)) %>%
  bsae::sf_to_nb()

@@

area_merged %>%
  filter(area_level == max(area_level)) %>%
  print(n = Inf)

neighbours_pairs_plot("log_or_gamma", 5) #' Nkhata Bay and neighbours
neighbours_pairs_plot("log_or_gamma", 26) #' Blantyre and neighbours
```

# NUTS specific assessment

```{r}
np <- bayesplot::nuts_params(mcmc)
```

Are there any divergent transitions?

```{r}
np %>%
  filter(Parameter == "divergent__") %>%
  summarise(n_divergent = sum(Value))

bayesplot::mcmc_nuts_divergence(np, bayesplot::log_posterior(mcmc))
```

We can also use energy plots (Betancourt 2017): ideally these two histograms would be the same 
When the histograms are quite different, it may suggest the chains are not fully exploring the tails of the target distribution.

```{r}
bayesplot::mcmc_nuts_energy(np)
```
